### 2019-07-11 manuscript/methods.tex
updated the figure of all models
Changed the layout and re-numbered the models accordingly.

The figure is now Fig S1.  It won't fit on a normal-size page, so it'll
have to be a separate supp file.  For now, the Makefile attaches the
giant figure as the final page of main.pdf.
edits to Models section of Methods
Relying now on the giant figure of all models.  The text isn't much
shorter, but I think its logic is clearer.

We employed a suite of SSE-style models \citep[BiSSE, MuSSE, HiSSE;][]{maddison_2007, fitzjohn_2012, beaulieu_2016} to test associations between ploidy, breeding system, and lineage diversification.
The full set of models is listed in \cref{table:marginallike}.

@@

In order to investigate the association between ploidy level and diversification, we first defined a binary state speciation and extinction model \citep[BiSSE,][]{maddison_2007} in which taxa were classified as diploid $(D)$ or polyploid $(P)$.
In this $D/P$ ploidy model, the model parameters are speciation rates ($\lambda_D$, $\lambda_P$) and extinction rates ($\mu_D$, $\mu_P$) for each state, and a rate of transition from D to P ($\rho$).
This is the same model \citet{mayrose_2011} used for this question across an array of plant genera.
% R- Itay wanted to hightlight that is the same model. Definitely a BiSSE. He did the ClaSSE later in the 2014 response.

% E: I moved r and nu to later, under statistical inference.  It made it seem like there were too many parameters here.
% I also removed italics around the model names because we don't actually use those names later on (more just the state codes).

To address diversification due to ploidy differences while also parsing out heterogeneity of diversification rates due another factor, we used models that include a hidden state \citep[HiSSE,][]{beaulieu_2016}. %B: inserted "(trait)", because that seems to be the goal - a biological explanation?  E: I would say hisse captures some factor that evolves like a binary trait, not necessarily a real biological trait
In these models, the observed diploid and polyploid states were each subdivided by a binary hidden trait with states $A$ and $B$.
In these $D/P+A/B$ ploidy and hidden state models, parameters include speciation rates ($\lambda_{D_A},\ \lambda_{D_B},\ \lambda_{P_A},\ \lambda_{P_B}$) and extinction rates ($\mu_{D_A},\ \mu_{D_B},\ \mu_{P_A},\ \mu_{P_B}$) for each subdivided state.
In the simplest case, the polyploidization rate, $\rho$, is the same within each hidden state, and the rates of transition between hidden states are symmetrical with parameter $\alpha$.
We also considered the more complex cases with different polyploidization rates in each hidden state ($\rho_A$ and $\rho_B$) and asymmetric transitions between hidden states ($\alpha$ and $\beta$).

% E: Rosana, please check this paragraph.  Did I get it correct?- R: Yes that is correct
To test if diversification rate heterogeneity is due only to a hidden trait and not the ploidy trait, we used character-independent diversification models \citep[CID,][]{beaulieu_2016}.
Parameters in this case are speciation and extinction rates for the hidden states ($\lambda_A,\ \lambda_B,\ \mu_A,\ \mu_B$), along with polyploidization ($\rho$) and transitions between hidden states ($(\alpha,\ \beta$).

Diagrams of these ploidy models are shown in \cref{figure:netdivall}AB.  % request from referee 2
% FIXME: ref to additional D/P models

@@

To assess the effects of breeding system in the diversification process, we first fit a model with self-incompatible $(I)$ or self-compatible $(C)$ states.
This is the same as the analysis of \citet{goldberg_2010} but with an updated phylogeny \citep{sarkinen_2013}.
This $I/C$ breeding system model has parameters for state-specific speciation ($\lambda_I,\ \lambda_C$) and extinction ($\mu_I,\ \mu_C$), and for transitions from $I$ to $C$ ($q_{IC}$).

To parse out the effect of breeding system on diversification while allowing for the possibility of heterogeneous diversification rates unrelated to breeding system, we subdivided each of the breeding system states into hidden states $A$ and $B$.
This $I/C+A/B$ breeding system and hidden state model has parameters for state-specific speciation ($\lambda_{I_A},\ \lambda_{I_B},\ \lambda_{C_A},\ \lambda_{C_B}$) and extinction ($\mu_{I_A},\ \mu_{I_B},\ \mu_{C_A},\ \mu_{C_B}$), for loss of SI ($q_{IC}$), and for transitions between hidden states ($\alpha$).
We also considered more complex models with different transition rates in each hidden state and asymmetric transitions between hidden states ($\alpha$ and $\beta$). % FIXME
Similar models were used by \citet{freyman_2017} used to study diversification in Onagraceae.

% E: Rosana, please check this paragraph.  Did I get it correct?
We further used character-independent diversification models to test if diversification rate heterogeneity is due only to a hidden trait and not the breeding system trait.
Parameters in this case are speciation and extinction rates for the hidden states ($\lambda_A,\ \lambda_B,\ \mu_A,\ \mu_B$), along with loss of SI ($q_{IC}$) and transitions between hidden states ($\alpha$).

@@

Diagrams of these breeding system models are shown in \cref{figure:netdivall}CD.  % request from referee 2
% FIXME: ref to additional I/C models

@@

Ploidy and breeding system might influence lineage diversification individually, but these two traits also have an intricate association as discussed in the Introduction.
Therefore, we considered a multi-state model that investigates the contribution of both traits and the allowable transitions between them.
This $ID/CD/CP$ ploidy and breeding system model has states for self-incompatible diploids $(ID)$, self-compatible diploids $(CD)$, and self-compatible polyploids $(CP)$; self-incompatible polyploids do not occur.
There are six parameters for state-specific diversification rates ($\lambda_{ID},\ \lambda_{CD},\ \lambda_{CP}$ for speciation; $\mu_{ID},\ \mu_{CD},\ \mu_{CP}$ for extinction) and three for transitions between states ($\rho_I,\ \rho_C$ for polyploidization transitions from $ID$ and $CD$ to $CP$, respectively; $q_{IC}$ for loss of self-incompatibility without polyploidization, from $ID$ to $CD$).
The total rate of loss of self-incompatibility, \ie transitions out of $ID$, is $q_{IC} + \rho_I$.
% E: Can we call it something other than q_IC?  It's a bit confusing to have the same parameter mean different things in different models.

% E: This paragraph is new.  Rosana, please check.
When including a new trait the question on whether 
the model with both breeding system and ploidy could provide a significantly better fit than the model with only ploidy was key.
Under all the statistical model comparison approaches, it is not valid to compare the $IC/CD/CP$ and $D/P$ model fits, as defined above, because they use different coding (i.e. (0,1) versus (0,1,2)). 
Similarly, the $IC/CD/CP$ and $I/C$ models cannot be compared statistically.
However, the evidence in the species and the tree are the same for both two-state and three-state models.
Instead, we define  and extend lumped models, as described first by \citet{tarasov_2019} for ontologies.  The lumped models proposed use the state space of the $IC/CD/CP$ model but constrain the rate parameters to mimic the behavior of the single-trait models. % should ref a figure

% E: We now have so many models to describe, I'm feeling like all this text may not be the best way.  I'm wondering if we can rely on figures/tables instead, and only write out the overall strategy. R- I feel the same but it might be very difficult to change at this point

Although the $ID/CD/CP$ model allows for two traits to potentially influence diversification, other unknown factors could as well.
We therefore added a hidden trait layer on top of our three-state model \citep[analogous to][]{caetano_2018, huang_2018}.
We fitted then $ID/CD/CP+A/B$ models, one with equal hidden rate $\alpha$ and equal transition rates between observed states to be independent of the hidden state (rates $\rho_I,\rho_C,\ q_{IC}$ as defined for the $ID/CD/CP$ model); second model were the hidden states are asymmetrical, and a third model were all parameters were different. For each of these models, there are additionally twelve diversification rate parameters ($\lambda_{ID_A},  \lambda_{ID_B},  \lambda_{CD_A}, \lambda_{CD_B}, \lambda_{CP_A}, \lambda_{CP_B}, \mu_{ID_A}, \mu_{ID_B}, \mu_{CD_A}, \mu_{CD_B},  \mu_{CP_A}, \mu_{CP_B}$).

Diagrams of these ploidy and breeding system models are shown in \cref{figure:netdivall}EF.  % request from referee 2
% FIXME: ref to additional models?

@@

Considering ploidy and breeding system together, there are two evolutionary pathways from SI diploid to SC polyploid \citep{brunet2001, robertson_2011}.
In the one-step pathway, the CP state is produced directly from the ID state when whole genome duplication disables SI.
In the two-step pathway, the CD state is an intermediate: SI is first lost, and later the SC diploid undergoes polyploidization.
We quantify the relative contribution of these pathways to polyploidy in two ways, each using the maximum \textit{a posteriori} (MAP) estimates of rates from the $IC/CD/CP$ model.

@@

\subsubsection{Diploidization as an exploratory hypothesis}

The models described above allow polyploidization but not its reverse.
We explored the effects of allowing diploidization by adding a parameter $\delta$ for transitions from $P$ back to $D$ (in the $D/P$ and $D/P+A/B$ models) or from $CP$ back to $CD$ (in the $I/C$ and $I/C+A/B$ models); we did not allow transitions from $CP$ back to $ID$ because that would include regain of SI.
Previous modeling approaches \citep{mayrose_2011} have argued against inferring diploidization rates when using ploidy data that comes from classifications based on chromosome number multiplicity or chromosome number change models like chromEvol \citep{mayrose_2010, glick2014, mayrose_2015, freyman_2017}, which assume only aneuploid change and polyploidization.
Our ploidy classifications, however, when possible also incorporated information on chromosome multiplicity at the genus level and author conclusions within original studies with alternative sources of information (\eg geographic distribution, genus ploidy distribution).
Since it is not known whether diploidization can be detected under these additional ploidy classifications, we fit models with and without diploidization in order to test whether the conclusions about diversification are sensitive to including diploidization.
% As discussed by \citet{servedio_2014}, the presence or absence of a process can have an exploratory goal.
In our models, the diploidization parameter (or its absence, $\delta=0$), represents an opportunity to explore a possibly significant assumption that may not decisively affect interactions among ploidy, breeding system, and diversification.

@@

\begin{table}
\begin{tabular}{lcccccc} \toprule
% \begin{tabular}{@{}lcccccc@{}} \toprule
%\multicolumn{4}{r}{Models} \\ \cmidrule(r){3-5}
%Model                     & Ploidy & Diploidization & Breeding  & Hidden & Num        & Marginal       \\
%                          &        &                & System    & State  & Parameters & Log-Likelihood \\ \midrule
%1. D/P                    & Yes    & No             & No        & No     & 5          & -1193.66\\
%2. D/P +$\delta$          & Yes    & Yes            & No        & No     & 6          & -1182.93 \\
%3. D/P+A/B                & Yes    & No             & No        & Yes    & 10         & -1150.99\\
%4. D/P+A/B+$\delta$       & Yes    & Yes            & No        & Yes    & 11         & \textbf{-1145.69}\\
%5. I/C                    & No     & No             & Yes       & No     & 5          & -1194.80 \\
%6. I/C+A/B                & No     & No             & Yes       & Yes    & 10         & \textbf{-1155.37}\\
%7. ID/CD/CP               & Yes    & No             & Yes       & No     & 9          & -1345.87\\
%8. ID/CD/CP+$\delta$      & Yes    & Yes            & Yes       & No     & 10         & -1344.50\\
%9. ID/CD/CP+A/B           & Yes    & No             & Yes       & Yes    & 15         & -1303.55 \\ 
%10. ID/CD/CP+A/B+$\delta$ & Yes    & Yes            & Yes       & Yes    & 16         & \textbf{-1300.35} \\
%\bottomrule
%\end{tabular}
%\caption{
   % The ten models and their marginal likelihoods.
    %Values in bold are for the best models within each class that are comparable (see \cref{table:bayesfactors}).
    %Abbreviations are D: diploid, P: polyploid, I: self-incompatible, C: self-compatible, A: one state of hidden trait, B: other state of hidden trait, $\delta$: diploidization.}
%\label{table:marginallike}
%\end{table}

>>>>

To test our hypotheses about lineage diversification and trait macroevolution, we fit a large suite of phylogenetic models.
All models are from the SSE framework \citep[BiSSE, MuSSE, HiSSE;][]{maddison_2007, fitzjohn_2012, beaulieu_2016}.
This style of model has parameters that describe per-lineage rates of speciation and extinction, specific to each character state (denoted $\lambda$ and $\mu$, respectively, with subscripts to indicate the state), along with rates of transitions between states (denoted $\rho$ for polyploidization, $\delta$ for diploidization, and $q_{IC}$ for loss of self-incompatibility).

The full set of models and all their rate parameters are detailed in \cref{fig:allmodels}.
Here we summarize how each type of model allows us to assess whether diversification is best explained by ploidy, breeding system, their combination, or some unknown factor.

@@

In order to investigate the association between ploidy level and diversification, we first defined a model in which taxa were classified as diploid $(D)$ or polyploid $(P)$ (M1).
This is the same model \citet{mayrose_2011} used for this question across an array of plant genera.
% R- Itay wanted to hightlight that is the same model. Definitely a BiSSE. He did the ClaSSE later in the 2014 response.
Although this model can be a powerful means to distinguish state-specific speciation and extinction from trait changes, it is prone to incorrectly reporting that a trait is associated with diversification differences \citep{maddison_2015, rabosky_2015}.
We therefore define several models that incorporate additional forms of diversification rate heterogeneity.

Our second ploidy model includes a binary hidden trait that subdivides each observed state (M2).
In this CID model \citep{beaulieu_2016}, the hidden trait can affect diversification but the observed trait does not.
Comparing M1 and M2 allows us to test whether diversification rate heterogeneity is better explained by ploidy or by some unknown factor.

We additionally fit three models in which both ploidy and a hidden trait could influence diversification (M3--M5).
These models differ in whether transitions between the hidden states are symmetric or asymmetric, and whether the polyploidization rate depends on the hidden state.
Comparing M1--M5 allows us to test whether ploidy is associated with diversification differences on top of differences explained by an unknown factor.

We further fit the analogues of these five models but including a rate parameter for transitions from polyploid to diploid (M6--M10).
This allows us to assess whether our conclusions about ploidy and diversification are robust to the possibility of diploidization.

@@

Our first five breeding system models follow the same logic as the first ploidy models.
We begin with each species classified as self-incompatible $(I)$ or self-compatible $(C)$ (M11).
This is the same as the analysis of \citet{goldberg_2010} but with an updated phylogeny \citep{sarkinen_2013}.

We then add models to allow diversification to be influenced by only a hidden trait (M12), or by both breeding system and a hidden trait (M13--M15, with varying degrees of complexity in the hidden trait transitions).
Similar models were used by \citet{freyman_2017} used to study diversification in Onagraceae.

@@

@@

Ploidy and breeding system might influence lineage diversification individually, but these two traits also have an intricate association (discussed in the Introduction).
Therefore, we considered several multi-state models that investigate the contribution of both traits and the allowable transitions between them.

The simplest classifies each species as either self-incompatible diploid $(ID)$, self-compatible diploid $(CD)$, or self-compatible polyploid $(CP)$; recall that self-incompatible polyploids do not occur (M16).
Each of these states may again be associated with different rates of speciation and extinction, and the allowable transitions are loss of SI within the diploid state (from $ID$ to $CD$), loss of SI via polyploidization (from $ID$ to $CP$), and polyploidization while SC (from $CD$ to $CP$).
% The total rate of loss of self-incompatibility, \ie transitions out of $ID$, is $q_{IC} + \rho_I$.
% E: Can we call it something other than q_IC?  It's a bit confusing to have the same parameter mean different things in different models.

As for the previous models of only one trait, we then allow diversification to be influenced by only a hidden trait (M17), or by both ploidy and breeding system as well as a hidden trait (M18--M20, with varying degrees of complexity in the hidden trait transitions).
We add this hidden trait layer on top of our three-state model \citep[similar to][]{caetano_2018, huang_2018}.
We also fit the analogous models but allowing for diploidization (M21--M25).

\subsubsection{Lumped models}

The models above allow us to assess the contributions of our two focal traits---ploidy and breeding system---to lineage diversification.
They do not, however, reveal whether it was ``worth it'' to include both traits in the analysis.
That is, they do not allow us to perform statistical model comparisons to test whether diversification heterogeneity is better explain by both traits than by either one of them.
This is because the models in the three sections above use different state spaces: $D/P$ versus $I/C$ versus $IC/CD/CP$.
Model comparisons work with the likelihood of the data given the model, but they are not valid when the data used differ from model to model.

In order to compare model fits for ploidy only versus breeding system only versus both traits, we use the technique of \citet{tarasov_2019} for lumping states together.
We use the state space of the $IC/CD/CP$ model but constrain the rate parameters to mimic the behavior of the single-trait models.
First we lump together $ID$ and $CD$ to form the diploid state, mimicking the $D/P$ model (M26).
Then we lump together $CD$ and $CP$ to form the self-compatible state, mimicking the $I/C$ model (M28).
We further add a hidden trait to each of these models (M27 and M29).

We do not include additional models with diploidization because this reverse transition renders the models non-lumpable \citep{tarasov_2019}.  % E: further explanation needed?

@@

Considering ploidy and breeding system together, there are two evolutionary pathways from SI diploid to SC polyploid \citep{brunet2001, robertson_2011}.
In the one-step pathway, the $CP$ state is produced directly from the $ID$ state when whole genome duplication disables SI.
In the two-step pathway, the $CD$ state is an intermediate: SI is first lost, and later the SC diploid undergoes polyploidization.
We quantify the relative contribution of these pathways to polyploidy in two ways, each using the maximum \textit{a posteriori} (MAP) estimates of rates from the simplest model that includes both traits (M16).

@@

% E: I don't think we need this section now that diploidization is in each of the sections above.  But I'm saving this text in case we want to put some of it in Results.
%--------------------------------------------------
% \subsubsection{Diploidization as an exploratory hypothesis}
%
% The models described above allow polyploidization but not its reverse.
% We explored the effects of allowing diploidization by adding a parameter $\delta$ for transitions from $P$ back to $D$ (in the $D/P$ and $D/P+A/B$ models) or from $CP$ back to $CD$ (in the $I/C$ and $I/C+A/B$ models); we did not allow transitions from $CP$ back to $ID$ because that would include regain of SI.
% Previous modeling approaches \citep{mayrose_2011} have argued against inferring diploidization rates when using ploidy data that comes from classifications based on chromosome number multiplicity or chromosome number change models like chromEvol \citep{mayrose_2010, glick2014, mayrose_2015, freyman_2017}, which assume only aneuploid change and polyploidization.
% Our ploidy classifications, however, when possible also incorporated information on chromosome multiplicity at the genus level and author conclusions within original studies with alternative sources of information (\eg geographic distribution, genus ploidy distribution).
% Since it is not known whether diploidization can be detected under these additional ploidy classifications, we fit models with and without diploidization in order to test whether the conclusions about diversification are sensitive to including diploidization.
% In our models, the diploidization parameter (or its absence, $\delta=0$), represents an opportunity to explore a possibly significant assumption that may not decisively affect interactions among ploidy, breeding system, and diversification.
%--------------------------------------------------

@@

\begin{suppfigure}
    \caption{ All models.  [This figure is provided as a separate, large-format page.] }
    \label{fig:allmodels}
\end{suppfigure}

%%% Below here is old tables.  I think the new versions should instead go in results.tex or a separate file. %%%

% \begin{table}
% \begin{tabular}{lcccccc} \toprule
% \begin{tabular}{@{}lcccccc@{}} \toprule
%\multicolumn{4}{r}{Models} \\ \cmidrule(r){3-5}
%Model                     & Ploidy & Diploidization & Breeding  & Hidden & Num        & Marginal       \\
%                          &        &                & System    & State  & Parameters & Log-Likelihood \\ \midrule
%1. D/P                    & Yes    & No             & No        & No     & 5          & -1193.66\\
%2. D/P +$\delta$          & Yes    & Yes            & No        & No     & 6          & -1182.93 \\
%3. D/P+A/B                & Yes    & No             & No        & Yes    & 10         & -1150.99\\
%4. D/P+A/B+$\delta$       & Yes    & Yes            & No        & Yes    & 11         & \textbf{-1145.69}\\
%5. I/C                    & No     & No             & Yes       & No     & 5          & -1194.80 \\
%6. I/C+A/B                & No     & No             & Yes       & Yes    & 10         & \textbf{-1155.37}\\
%7. ID/CD/CP               & Yes    & No             & Yes       & No     & 9          & -1345.87\\
%8. ID/CD/CP+$\delta$      & Yes    & Yes            & Yes       & No     & 10         & -1344.50\\
%9. ID/CD/CP+A/B           & Yes    & No             & Yes       & Yes    & 15         & -1303.55 \\ 
%10. ID/CD/CP+A/B+$\delta$ & Yes    & Yes            & Yes       & Yes    & 16         & \textbf{-1300.35} \\
%\bottomrule
%\end{tabular}
%\caption{
   % The ten models and their marginal likelihoods.
    %Values in bold are for the best models within each class that are comparable (see \cref{table:bayesfactors}).
    %Abbreviations are D: diploid, P: polyploid, I: self-incompatible, C: self-compatible, A: one state of hidden trait, B: other state of hidden trait, $\delta$: diploidization.}
%\label{table:marginallike}
%\end{table}
